### 2025-09-24 paper/sec/evaluation.tex
Tweaks to eval.

\subsection{Validation}
\label{subsec:validation-with-counterfactual-data}

The validation process is structured in two phases

\paragraph{Initial validation}
In the initial phase, the generated expression is executed via Fluid CLI. If the execution is successful, the
result obtained is compared with the expected result. In case of error, the output is forwarded to the
\InterpretationAgent, which proceeds to regenerate the expression.

\paragraph{Validation via counterfactual testing}
In the second phase, the data used for execution is deliberately modified in order to verify the consistency
of the behaviours. Both the expected and generated expressions are executed again and the results are
compared. If discrepancies emerge, the error is forwarded to the \InterpretationAgent. If there are no
differences, the expression is considered valid and is returned to the user.

@@

These seem quite pertinent to the overall problem of transparent science communication:
\begin{itemize}
\item How does performance degrade with partial/ambiguous information?
\item How is performance affected by ``adversarial'' (intentionally/unintentionally misleading) information?
\end{itemize}

@@

\subsection{Success Rate by Category}
To evaluate the system we used a sample of the Scigen dataset~\citep{scigen_dataset_2021}.

We aggregated the generated expression in three categories: aggregation, trends, quantitative (See
Table~\ref{tab:natural-language-forms}). \figref{success_rate_by_category} shows the results obtained in
generating expression by categories.

@@

\subsection{Results}

\begin{table}[h!]
    \scriptsize

    \centering
    \begin{subtable}{0.45\linewidth}
        \centering
        \begin{tabular}{lcccc}
            \hline
            & \multicolumn{2}{c}{\textbf{gpt4o}} & \multicolumn{2}{c}{\textbf{gpt5}} \\
            & \textbf{Success Rate} & \textbf{SD} & \textbf{Success Rate} & \textbf{SD} \\
            \hline
            Comparison     & 74   & 3 & 77.4 & 2 \\
            Data Retrieval & 76   & 3 & 77.4 & 2 \\
            Difference     & 80   & 3 & 77.4 & 2 \\
            Min/max        & 50   & 3 & 77.4 & 2 \\
            Rank           & 44   & 3 & 77.4 & 2 \\
            Total          & 33   & 3 & 77.4 & 2 \\
            Overall        & 74.9 & 3 & 77.4 & 2 \\
            \hline
        \end{tabular}
        \caption{Results with target-value = true}
        \label{tab:results_true}
    \end{subtable}%
    \hfill
    \begin{subtable}{0.45\linewidth}
        \centering
        \begin{tabular}{lcccc}
            \hline
            & \multicolumn{2}{c}{\textbf{gpt4o}} & \multicolumn{2}{c}{\textbf{gpt5}} \\
            & \textbf{Success Rate} & \textbf{SD} & \textbf{Success Rate} & \textbf{SD} \\
            \hline
            Comparison     & -- & -- & -- & -- \\
            Data Retrieval & -- & -- & -- & -- \\
            Difference     & -- & -- & -- & -- \\
            Min/max        & -- & -- & -- & -- \\
            Rank           & -- & -- & -- & -- \\
            Total          & -- & -- & -- & -- \\
            Overall        & -- & -- & -- & -- \\
            \hline
        \end{tabular}
        \caption{Results without target-value}
        \label{tab:results_no_target}
    \end{subtable}
    \caption{Comparison of success rates (with standard deviation) between \textbf{gpt4o} and \textbf{gpt5} across different task categories, reported with and without the use of target-values.}
    \label{tab:results_combined}
\end{table}

In this section, we report the experimental results.
Table \ref{tab:results_combined} summarizes the performance of \textit{gpt4o} and the most recent OpenAI model, \textit{gpt5}, under two different settings: with target-value sharing enabled and without target-value sharing.
Overall, the system produced a valid fluid expression in 74.9\% of the cases (S.D. 3\%).

Figure~\ref{fig:success_rate_by_category} illustrates the success rate for each category, both with and without target-value sharing.
The results show that providing the target value is crucial for achieving a high success rate.
Moreover, the \InterpretationAgent generates a correct expression in approximately XX\% of the cases.

The system achieved performance above 70\% in comparison, data retrieval, and min/max search tasks.
In contrast, accuracy decreased significantly in expressions requiring differences (20\%) and in ranking tasks (50\%).

Figure~\ref{fig:success_rate_by_category_count} reports the success rate as a function of the number of categories assigned to each expression.
Fluid instructions combining multiple operations were associated with multiple categories.
Results show that the success rate was 75\% for simple expressions (one category), increased to 85\% for two categories,
and dropped to 58\% when three categories were involved.

In addition to evaluating performance when the target expression is provided, we also assessed the model’s ability to produce a
correct expression when no reference string was given.
On the SciGen dataset, the model achieved a 74.9\% success rate when provided with the target string.
Without the target string, performance decreased moderately to 39.4\%, indicating that the model retains substantial
ability to reconstruct valid solutions even without explicit guidance

\subsection{Problematic Cases}
\label{subsec:problematic-cases}

Table~\ref{tab:problematic_cases} presents examples of problematic cases in which the system was asked to generate a valid
fluid expression.
In these cases, producing a valid expression is challenging because the input data (or the NL query) contains ambiguities
that prevent the construction of a well-formed expression.

For instance, in the \textit{missing-data} problem, the \InterpretationAgent should generate a function to compute
 the average of the emissions. However, the input data does not provide information for all months.

In such cases, the system was unable to generate a general expression and instead produced the expected value directly.

@@

        \hline
    \end{tabular}
    \caption{Categories of problematic example}
    \label{tab:problematic_cases}
\end{table}

>>>>

@@

\begin{itemize}
\item How does performance degrade with partial/ambiguous information?
\item How is performance affected by ``adversarial'' (intentionally/unintentionally misleading) information?
\end{itemize}

@@

\subsection{Success Rate by Category}
To evaluate the system we used a sample of the Scigen dataset~\citep{moosavi21}. We aggregated the generated
expression in three categories: aggregation, trends, quantitative (See
Table~\ref{tab:natural-language-forms}). \figref{success_rate_by_category} shows the results obtained in
generating expression by categories.

@@

\subsection{Counterfactual testing}
\label{subsec:validation-with-counterfactual-data}

For the \emph{counterfactual testing} part of our analysis, the raw data execution is deliberately modified in
order to verify the consistency of the behaviours. Both the expected and generated expressions are executed
again and the results are compared. If discrepancies are found \rpnote{complete}

\subsection{Negative Cases}
\label{subsec:negative-cases}

@@

        \hline
    \end{tabular}
    \caption{Categories of problematic example}
    \label{tab:ambiguita}
\end{table}
